Multilocus Genotype And Mycelial Compatability Group Assessment

### 2017-05-06 doc/RMD/mlg-mcg.Rmd
add tables and graphs

setPop(dat11) <- ~MLGRegion
pal <- unlist(map2(LETTERS[2:4], comm, function(a, b) setNames(viridis::viridis(length(b), option = a), b)))
pal <- c(pal, setNames(rep("#F6F6F6FF", 3), c("Mexico", "ID", "WI")))
```

@@

These tables are a good start, but we want to have a publication-ready table:

@@

psex_from_graph <- function(g, MLG){
  weights <- setNames(E(g)$weight, E(g)$label)
  weights <- split(weights, names(weights))[MLG]
  1 -vapply(weights, unique, numeric(1))
}

mlg_table <- mlgs %>%
  slice(1:5L) %>%
  mutate(MLG = as.character(MLG)) %>%
  mutate(Psex = psex_from_graph(graph11loc$total, MLG)) %>%
  unnest() %>%
  group_by(MLG, Samples, Psex) %>%
  summarize(MCG = paste(MCG, collapse = ", ")) %>%
  rename(N = Samples) %>%
  arrange(desc(N))

kable(mlg_table, caption = "Top 5 MLGs", format.args = list(digits = 3))

mcg_table <- mcgs %>%
  slice(1:5L) %>%
  mutate(MCG = as.character(MCG)) %>%
  unnest() %>%
  group_by(MCG, Samples, Evenness) %>%
  summarize(MLG = list(strwrap(paste(MLG, collapse = ", "), 28))) %>%
  rename(N = Samples) %>%
  arrange(desc(N)) %>% 
  unnest() %>%
  ungroup() %>%
  mutate(Evenness = ifelse(!duplicated(MCG), format(Evenness, digits = 3), "")) %>%
  mutate(N = ifelse(!duplicated(MCG), N, "")) %>%
  mutate(MCG = ifelse(!duplicated(MCG), MCG, ""))
kable(mcg_table, caption = "Top 5 MCGs")
```

@@

```{r barplots, fig.width = 10, fig.height = 4}
mcgs %>% 
  gather(type, count, MLGs, Samples, -Evenness) %>%
  arrange(desc(type), desc(count)) %>%
  mutate(MCG = forcats::fct_inorder(MCG, ordered = TRUE)) %>%
  ggplot(aes(x = MCG, y = count, group = type, fill = type, alpha = Evenness)) +
  geom_col(aes(width = ifelse(type == "MLGs", 0.5, 0.85)), color = "black", position = "identity") +
  annotate(geom = "text", x = 13, y = 51, label = sprintf("Mean Evenness: %.3f", mean(mcgs$Evenness, na.rm = TRUE))) +
  scale_fill_viridis(end = 0.75, discrete = TRUE, direction = -1) +
  scale_y_continuous(expand = c(0, 2)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  ggtitle("How Evenly are Multilocus Genotypes (MLGs) spread across MCGs?")
mlgs %>% 
  gather(type, count, MCGs, Samples, -Evenness) %>%
  arrange(desc(type), desc(count)) %>%
  mutate(MLG = forcats::fct_inorder(MLG, ordered = TRUE)) %>%
  ggplot(aes(x = MLG, y = count, group = type, fill = type)) +
  geom_col(aes(width = ifelse(type == "MCGs", 0.5, 0.85), alpha = Evenness), color = "black", position = "identity") +
  annotate(geom = "text", x = 20, y = 21, label = sprintf("Mean Evenness: %.3f", mean(mlgs$Evenness, na.rm = TRUE))) +
  scale_fill_viridis(end = 0.75, discrete = TRUE, direction = -1) +
  scale_y_continuous(expand = c(0, 2)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  ggtitle("How Evenly are MCGs spread across Multilocus Genotypes (MLGs)?")
```

@@

```{r severity, fig.width = 7, fig.height = 4}

strat <- bind_cols(strata(dat11), 
                   other(dat11)$meta, 
                   data_frame(MLG = mll(dat11, "original")))
count_group <- . %>% 
  mutate(nobs = n()) %>%  # count the number of samples/MCG
  ungroup() %>%
  arrange(desc(nobs)) # arrange by number of samples and reorder factors 
Severity <- filter(strat, MCG %in% mcgs$MCG[1:5]) %>%
  group_by(MCG) %>% 
  count_group %>%
  mutate(MCG = forcats::fct_inorder(factor(MCG), ordered = TRUE)) %>%
  mutate(MLG = ifelse(MLG %in% mlgs$MLG[1:5], paste("MLG", MLG), "Other"))

@@

Note in this case we are assigning MLGs to the communities indicated. This means
that even though they share a connection with an MCG in a different community,
they belong to the MCG community they were initially assigned to. We will also
assign communities based on what MCG they belong to.

```{r treebuilding, cache = TRUE}
the_communities
strat_mlg <- strat %>% 
  mutate(MLG = paste0("MLG.", MLG)) %>%
  left_join(the_communities, by = c("MLG" = "vertex"))
strat_mcg <- strat %>% 
  left_join(the_communities, by = c("MCG" = "vertex"))

datdist <- bruvo.dist(dat11, replen = other(dat)$REPLEN)
datree <- phangorn::upgma(datdist)
```

Now we can visualize that tree.

```{r, fig.width = 7, fig.height = 7}
ape::plot.phylo(ape::ladderize(datree), type = "fan", no.margin = FALSE, tip.col = viridis(4, direction = -1)[as.integer(factor(strat_mlg$comm))])
legend(-0.015, 0.11, legend = levels(factor(strat_mlg$comm)), fill = viridis(4, direction = -1), bty = "n")
title("MLG-based assignment")
ape::plot.phylo(ape::ladderize(datree), type = "fan", no.margin = FALSE, tip.col = viridis(4, direction = -1)[as.integer(factor(strat_mcg$comm))])
legend(-0.015, 0.11, legend = levels(factor(strat_mcg$comm)), fill = viridis(4, direction = -1), bty = "n")
title("MCG-based assignment")
```

From what I can see, it appears that there's not much of a tight correlation of 
genetic structure and MCG.

>>>>

setPop(dat11) <- ~MLGRegion
pal <- unlist(map2(LETTERS[2:4], comm, function(a, b) setNames(viridis::viridis(length(b), option = a), b)))
pal <- c(pal, setNames(rep("#F6F6F6FF", 3), c("Mexico", "ID", "WI")))

strat <- bind_cols(strata(dat11), 
                   other(dat11)$meta, 
                   data_frame(MLG = mll(dat11, "original")))
```

@@

These tables are a good start, but we want to have a publication-ready table.

@@

psex_from_graph <- function(g, MLG){
  weights <- setNames(E(g)$weight, E(g)$label)
  weights <- split(weights, names(weights))
  if (length(MLG) > 1){
    weights <- weights[MLG]
    res <- 1 - vapply(weights, unique, numeric(1))
  } else {
    res <- 1 - unique(weights[[MLG]])
  }
  res
}

# Paste region and count vectors
pasteRC <- function(R, C){
  ord <- rev(order(C))
  RC <- paste(R[ord], C[ord], sep = " (")
  RC <- paste(RC, collapse = "), ")
  paste0(RC, ")")
}

mlg_table <- mcgmlg %>%
  filter(MLG %in% mlgs$MLG[1:5]) %>%
  mutate(MLG = as.integer(MLG)) %>%
  select(-Freq) %>%
  inner_join(strat) %>%
  group_by(MLG, MCG, Region) %>%
  summarize(N = n()) %>%
  ungroup() %>%
  group_by(MLG, MCG) %>%
  summarize(Region = pasteRC(Region, N), N = sum(N)) %>%
  ungroup() %>%
  mutate(mlmatch = match(MLG, mlgs$MLG[1:5])) %>%
  arrange(mlmatch, desc(N), MCG) %>%
  group_by(MLG) %>%
  mutate(Psex = psex_from_graph(graph11loc$total, as.character(unique(MLG)))) %>%
  ungroup() %>%
  select(MLG, Psex, MCG, Region) %>%
  mutate(Psex = ifelse(duplicated(MLG), "", format(Psex, scientific = FALSE, digits = 2))) %>%
  mutate(MLG = ifelse(duplicated(MLG), "", as.character(MLG)))

mcg_table <- mcgmlg %>%
  filter(MCG %in% mcgs$MCG[1:5]) %>%
  mutate(MLG = as.integer(MLG)) %>%
  select(-Freq) %>%
  inner_join(strat) %>%
  group_by(MLG, MCG, Region) %>%
  summarize(N = n()) %>%
  ungroup() %>%
  group_by(MLG, MCG) %>%
  summarize(Region = pasteRC(Region, N), N = sum(N)) %>%
  ungroup() %>%
  mutate(mlmatch = match(MCG, mcgs$MCG[1:5])) %>%
  arrange(mlmatch, desc(N), MLG) %>%
  select(MCG, MLG, Region) %>%
  mutate(MCG = ifelse(duplicated(MCG), "", as.character(MCG))) %>%
  mutate(MLG = as.character(MLG))

kable(mlg_table, caption = "Top 5 MLGs with associated MCGs and Regions (number of samples in parenthesis)")
if (!dir.exists(file.path(PROJHOME, "results/tables"))) dir.create(file.path(PROJHOME, "results/tables"))
readr::write_csv(mlg_table, path = file.path(PROJHOME, "results/tables/mlg_table.csv"))

kable(mcg_table, caption = "Top 5 MCGs with associated MLGs and Regions (number of samples in parenthesis)")
if (!dir.exists(file.path(PROJHOME, "results/tables"))) dir.create(file.path(PROJHOME, "results/tables"))
readr::write_csv(mcg_table, path = file.path(PROJHOME, "results/tables/mcg_table.csv"))
```

@@

```{r barplots, fig.width = 10, fig.height = 4}
mcgs %>% 
  gather(type, count, MLGs, Samples, -Evenness) %>%
  arrange(desc(type), desc(count)) %>%
  mutate(MCG = forcats::fct_inorder(MCG, ordered = TRUE)) %>%
  ggplot(aes(x = MCG, y = count, group = type, fill = type, alpha = Evenness)) +
  geom_col(aes(width = ifelse(type == "MLGs", 0.5, 0.85)), color = "black", position = "identity") +
  annotate(geom = "text", x = 13, y = 51, label = sprintf("Mean Evenness: %.3f", mean(mcgs$Evenness, na.rm = TRUE))) +
  # scale_fill_viridis(end = 0.75, discrete = TRUE, direction = -1) +
  scale_fill_manual(values = c("black", "white")) +
  scale_y_continuous(expand = c(0, 2)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  ggtitle("How Evenly are Multilocus Genotypes (MLGs) spread across MCGs?")
mlgs %>% 
  gather(type, count, MCGs, Samples, -Evenness) %>%
  arrange(desc(type), desc(count)) %>%
  mutate(MLG = forcats::fct_inorder(MLG, ordered = TRUE)) %>%
  ggplot(aes(x = MLG, y = count, group = type, fill = type)) +
  geom_col(aes(width = ifelse(type == "MCGs", 0.5, 0.85), alpha = Evenness), color = "black", position = "identity") +
  annotate(geom = "text", x = 20, y = 21, label = sprintf("Mean Evenness: %.3f", mean(mlgs$Evenness, na.rm = TRUE))) +
  # scale_fill_viridis(end = 0.75, discrete = TRUE, direction = -1) +
  scale_fill_manual(values = c("black", "white")) +
  scale_y_continuous(expand = c(0, 2)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  ggtitle("How Evenly are MCGs spread across Multilocus Genotypes (MLGs)?")
```

@@

```{r severity, fig.width = 7, fig.height = 4}
count_group <- . %>% 
  mutate(nobs = n()) %>%  # count the number of samples/MCG
  ungroup() %>%
  arrange(desc(nobs)) # arrange by number of samples and reorder factors 
Severity <- filter(strat, MCG %in% mcgs$MCG[1:5]) %>%
  group_by(MCG) %>% 
  count_group %>%
  mutate(MCG = forcats::fct_inorder(factor(MCG), ordered = TRUE)) %>%
  mutate(MLG = ifelse(MLG %in% mlgs$MLG[1:5], paste("MLG", MLG), "Other"))

@@

```{r treebuilding, cache = TRUE}
strat_mcg <- left_join(strat, the_communities, by = c("MCG" = "vertex"))
datdist   <- bruvo.dist(dat11, replen = other(dat)$REPLEN)
datree    <- phangorn::upgma(datdist)
```

Now we can visualize that trees.

```{r trees, fig.width = 7, fig.height = 7}
ape::plot.phylo(ape::ladderize(datree), type = "fan", no.margin = FALSE, tip.col = viridis(4, direction = -1)[as.integer(factor(strat_mcg$comm))])
legend(-0.015, 0.11, legend = levels(factor(strat_mcg$comm)), fill = viridis(4, direction = -1), bty = "n")
title("MCG-based assignment")

ape::plot.phylo(ape::ladderize(datree), type = "fan", no.margin = FALSE, tip.col = plasma(366)[rank(strat$Severity, ties.method = "first")])
legend(-0.015, 0.125, legend = quantile(strat$Severity), fill = plasma(5), bty = "n")
title("Severity")
```

From what I can see, it appears that there's not much of a tight correlation of 
genetic structure and MCG. There is the clear cluster of MLG 66 in Community C,
but there are other isolates in distant parts of the tree. I'm wondering what
happens if we average the distances between groups.

```{r group_dist}
datmat <- as.matrix(datdist)
distmcg <- strat_mcg %>% 
  group_by(comm) %>%
  summarize(dist = list(as.dist(datmat[Isolate, Isolate])), N = n())
distmcg %>%
  group_by(comm) %>%
  mutate(meandist = mean(dist[[1]]),
         mediandist = median(dist[[1]])) %>%
  ungroup() %>%
  select(comm, N, meandist, mediandist)
distmcg %>% 
  unnest() %>%
  ggplot(aes(x = dist, fill = comm)) +
  geom_density(alpha = 0.5) +
  geom_rug(aes(color = comm), alpha = 0.12) +
  scale_fill_viridis(direction = -1, discrete = TRUE) +
  scale_color_viridis(direction = -1, discrete = TRUE) +
  ggtitle("Bruvo's distance by MCG community") +
  theme_bw(base_size = 14)
```

This is quite distant. For reference, a distance of `r round((0.5/11)*7, 2)` is
on average 7 steps.
